Hierarchical models and seq model fits

### 2025-04-03 documentation/hierarchicalmodels_seqruns.qmd
document hierarchical data functions

The set up and notation in stan/R, in terms of parameters, using parameter Omega as an example, is currently as follows:

-   Omega_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.

-   Omega_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the original package, it was called Omega_sigma_raw).

-   Omega_star refers to the vector with (mu global and ALL $\eta$s). we get Omega_stars from Omega_raw and the vector ($s_{\mu, global}$ , Omega_sigma), using the stan `scale_blocks` function.

-   Omega refers to the final $\mu$s at the country level. Matrix multiplication is used to obtain the Omega country vector from the Omega_stars. The full matrix `Omega_model_matrix` is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas). The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of `Omega_model_matrix` into different parts U, V, W.

-   Because of the sequential fitting, for Omega_raw and Omega_sigma, we also had to distinguish between parameters that were fixed vs those that were estimated. We do so with the \_fixed and \_estimate additions, i.e. the vector of Omega_raws is produced from Omega_raw_fixed and Omega_raw_estimate.

@@

in the simpler model used, various Omegas are defined in the transformed parameters block so we can check those directly. (Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).

here just checking the sigmas and first Omega:

@@

fit <- fit1a
 parname_list <- list( c("Omega_raw_estimate[1]",
                          "Omega_sigma_estimate"))
#    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
#    pdf(plot_dirandname, width = 11, height = 11)
    for (i in 1:length(parname_list)){
      parname <- parname_list[[i]]
      samp <- fit$samples$draws(parname )
      #mcmc_trace(samp)
      p <- mcmc_dens_overlay(samp)
      # for more than 1 parameter
      #mcmc_pairs(samp)
      print(p)
    }
#    dev.off()
```

@@

-   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the Omega_stars, that's what the original helper functions use as inputs.

-   We clean up helper functions, so that they take in the Omega_stars, and specific other stan_data, and are better documented. See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.

```{r}
fit <- fit1a
# some notes, to understand updates between this code base and original one
# and where to find info 
#fit1a$country_index is now fit$geo_unit
# see fit1a$hierarchical_level for info
# hierarchical_data(fit$geo_unit, fit$hierarchical_level)
# note that subhierarchies are called column in hierarchical data
Omega <- list()
for(subhierarchy in fit$hierarchical_level) {
  Omega[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "Omega",
      fit_samples = fit$samples)
}
# Omega[[]] has posterior samples for the subhierarchy level
# to get the summaries
quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
  tibble(
    val = quantile(x, probs, na.rm = TRUE),
    quant = probs
  )
}
res_global <- map(Omega, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_global

@@

```{r}
fit <- fit_local
Omega <- list()
for(subhierarchy in fit$hierarchical_level) {
  Omega[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "Omega",
      fit_samples = fit$samples)
}
res_local <- map(Omega, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_local
```

@@

-   How to get the Omegas in full model set up: in the simpler model used here, various Omegas are defined in the transformed parameters block so we can check those directly. Note that this not (yet) true for the full transition model because the creation of parameters takes place within the process model function. Options to obtain the Omegas in the full model are to

    1.  add them to generated quantities: pull out omegas with an `if {add_output_omegas}` type statement, re-using functions/statements introduced in the simpler file here.

    2.  rewrite transformed parameter block and pull Omegas out of the process model function (not sure if that slows things down when not used).

>>>>

The set up and notation in stan/R, in terms of parameters, using parameter mu as an example, is currently as follows:

-   mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.

-   mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the original package, it was called mu_sigma_raw).

-   mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw and the vector ($s_{\mu, global}$ , mu_sigma), using the stan `scale_blocks` function.

-   mu refers to the final $\mu$s at the country level. Matrix multiplication is used to obtain the mu country vector from the mu_stars. The full matrix `mu_model_matrix` is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas). The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of `mu_model_matrix` into different parts U, V, W.

-   Because of the sequential fitting, for mu_raw and mu_sigma, we also had to distinguish between parameters that were fixed vs those that were estimated. We do so with the \_fixed and \_estimate additions, i.e. the vector of mu_raws is produced from mu_raw_fixed and mu_raw_estimate.

@@

in the simpler model used, various mus are defined in the transformed parameters block so we can check those directly. (Note that this not (yet) true for the transition model because the creation of parameters takes place within the transition function, see notes in last section on this).

here just checking the sigmas and first mu:

@@

fit <- fit1a
 parname_list <- list( c("mu_raw_estimate[1]",
                          "mu_sigma_estimate"))
#    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
#    pdf(plot_dirandname, width = 11, height = 11)
    for (i in 1:length(parname_list)){
      parname <- parname_list[[i]]
      samp <- fit$samples$draws(parname )
      #mcmc_trace(samp)
      p <- mcmc_dens_overlay(samp)
      # for more than 1 parameter
      #mcmc_pairs(samp)
      print(p)
    }
#    dev.off()
```

@@

-   To deal with the set up where things are fixed vs estimated, we let the stan model spit out the mu_stars, that's what the original helper functions use as inputs.

-   We clean up helper functions, so that they take in the mu_stars, and specific other stan_data, and are better documented. See function extract_parameter_subhierarchical_simplified in fit_model_simplified R script, and its usage below.

```{r}
fit <- fit1a
# some notes, to understand updates between this code base and original one
# and where to find info 
#fit1a$country_index is now fit$geo_unit
# see fit1a$hierarchical_level for info
# hierarchical_data(fit$geo_unit, fit$hierarchical_level)
# note that subhierarchies are called column in hierarchical data
mu <- list()
for(subhierarchy in fit$hierarchical_level) {
  mu[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "mu",
      fit_samples = fit$samples)
}
# mu[[]] has posterior samples for the subhierarchy level
# to get the summaries
quantile_df <- function(x, probs = c(0.025, 0.5, 0.975)) {
  tibble(
    val = quantile(x, probs, na.rm = TRUE),
    quant = probs
  )
}
res_global <- map(mu, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_global

@@

```{r}
fit <- fit_local
mu <- list()
for(subhierarchy in fit$hierarchical_level) {
  mu[[subhierarchy]] <-
    extract_parameter_subhierarchical_simplified (
      hierarchical_data = hierarchical_data(fit$geo_unit, fit$hierarchical_level),
      subhierarchy = subhierarchy,
      parname = "mu",
      fit_samples = fit$samples)
}
res_local <- map(mu, function(tibble_samples) 
  tibble_samples %>% select(name, value)  %>% reframe(quantile_df(value), .by = name))
res_local
```

@@

-   How to get the mus in full model set up: in the simpler model used here, various mus are defined in the transformed parameters block so we can check those directly. Note that this not (yet) true for the full transition model because the creation of parameters takes place within the process model function. Options to obtain the mus in the full model are to

    1.  add them to generated quantities: pull out mus with an `if {add_output_mus}` type statement, re-using functions/statements introduced in the simpler file here.

    2.  rewrite transformed parameter block and pull mus out of the process model function (not sure if that slows things down when not used).
